---
title: "Rolling Stone Data Analysis"
authors: "Julia Czekańska, Anastasiia Vudvud"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
# Define the required libraries
libs <- c("tidyverse", "ggtext", "patchwork", "cowplot", "gridExtra", "lubridate", "ggthemes", "tm", "SnowballC", "wordcloud", "RColorBrewer", "RCurl", "XML", "tidytext", "viridis", "pals", "dplyr","ggplot2", "tidyr", "stringr", "gganimate", "ggraph", "tidygraph")

# Check if libraries are installed; install missing ones
installed_libs <- libs %in% rownames(installed.packages())

if (any(installed_libs == FALSE)) {
    install.packages(libs[!installed_libs])
} else {
    print("All the libraries already installed")
}

# Load the libraries
library(tidyverse)
library(tidytext)
library(ggtext)
library(patchwork)
library(cowplot)
library(gridExtra)
library(lubridate)
library(ggthemes)
library(tm)
library(SnowballC)
library(wordcloud)
library(RColorBrewer)
library(RCurl)
library(NLP)
library(XML)
library(pals)
library(gganimate) 
library(transformr) 
library(dplyr) 
library(ggplot2) 
library(tidyr)
library(stringr)
library(ggraph)
library(tidygraph) 
```

```{r}
df <- read.csv("rolling_stone.csv", stringsAsFactors = FALSE)
string_columns <- c("sort_name", "clean_name", "album", "genre", "type", "spotify_url", "artist_gender")

df[string_columns] <- df[string_columns] %>% lapply(str_to_title)
```

## EDA

### Gender Distribution among Artists

Artists were divided into groups and soloists and then amount of male, female, or mixed gender was counted.
The data shows that male artists dominate the rankings and tend to be in bands more often than female artists.

```{r}
gender_distribution <- df %>% 
  mutate(artist_type = ifelse(artist_member_count == 1, "Solo", "Group")) %>%
  count(artist_type, artist_gender) %>%
  filter(!is.na(artist_type)) %>% 
  spread(artist_type, n, fill = 0)
gender_distribution
```

### Artist Member Count vs Debut Album Release Year

The debut years of artists were grouped by decade, and the average number of band members per decade was calculated.
The results highlight the rise of bands in the '60s-'80s.

```{r}
member_vs_debut <- df %>% 
  mutate(debut_decade = floor(debut_album_release_year / 10) * 10) %>%
  group_by(debut_decade) %>%
  summarise(avg_member_count = round(mean(artist_member_count, na.rm = TRUE), 2)) %>% 
  filter(!is.na(debut_decade))
member_vs_debut
```

### Albums with More Than Eight Words in the Title

Album names with more than 8 words were selected, which resulted in 13 names.

```{r}
long_album_titles <- df %>% 
  mutate(word_count = str_count(album, " ") + 1, avg_rank = round(rowMeans(select(., rank_2003, rank_2012, rank_2020), na.rm = TRUE), 2)) %>% 
  filter(word_count > 8) %>% 
  select(album, clean_name, avg_rank) %>% 
  arrange(avg_rank)

long_album_titles
```

### Average Rank Based on Word Usage in Album Title

Album titles were tokenized into individual words, their frequency was counted, and the average rank was calculated for words appearing at least 10 times.
The most common words are generally neutral everyday words.

```{r}
word_rank <- df %>%
  separate_rows(album, sep = " ") %>%
  group_by(album) %>%
  summarise(appearance_count = n(), avg_rank = round(mean(c(rank_2003, rank_2012, rank_2020), na.rm = TRUE), 2)) %>%
  filter(appearance_count >= 10) %>%
  arrange(desc(appearance_count))

word_rank
```

### Top genres by average Spotify popularity

At first we need to remove rows where genre is NA and group by genre.
After doing this, we need to find average from column spotify_popularity.
The range is 0 - 100 and the buggest the number, more popular the album is.
So the most popular are Latin albums

```{r}
df %>%
  filter(!is.na(genre)) %>%
  group_by(genre) %>%
  summarise(avg_popularity = mean(spotify_popularity, na.rm = TRUE)) %>% 
  arrange(desc(avg_popularity))
```

### Average age of band members for each genre at the moment of debut

At first we need to calculate average age of artists at the moment of debut.
After that we filter the data, so that we only take bands, which has genre.
We group by genre and find average value.
Groups with the youngest members are in Rock N' Roll/Rhythm & Blues and the oldest in Latin.

```{r}
df %>%
  mutate(avg_age = (release_year - (artist_birth_year_sum / artist_member_count))) %>%
  filter(artist_member_count > 1 & !is.na(genre)) %>% 
  group_by(genre) %>%
  summarise(avg_band_age = mean(avg_age, na.rm = TRUE)) %>%
  arrange(desc(avg_band_age)) 
```

### Number of albums from each genre in each rank

To count the numbers of albums in each rank by genre we needed to transform columns like "rank_2003" to longer data.
Name of the column transformed to "rank_year" and values were added to column "rank".
After that we needed to filter out genres and rank, group by it and summarize.

```{r}
number_of_albums_from_each_genre_in_each_rank <- df %>%
    pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>%
  filter(!is.na(rank) & !is.na(genre)) %>%
  group_by(genre, rank_year) %>% 
  summarise(album_count = n(), .groups = "drop") 

number_of_albums_from_each_genre_in_each_rank
```

### Top 10 albums that improved the most (positive) and dropped the most (negative)

At first we need to filter the albums that aren't in 2003 or 2020 rankings.
After we arrange them by *differential* ascending or descending and choose top 10.

```{r}
top_improved <- df %>%
  filter(!is.na(rank_2003) & !is.na(rank_2020)) %>%
  arrange(desc(differential)) %>% 
  select(clean_name, album, rank_2003, rank_2020, differential) %>%
  slice_head(n = 10)
top_improved

top_dropped <- df %>%
  filter(!is.na(rank_2003) & !is.na(rank_2020)) %>%
  arrange(differential) %>%
  select(clean_name, album, rank_2003, rank_2020, differential) %>%
  slice_head(n = 10)
top_dropped
```

### Which percent of the albums in top charts have special characters in their name?

At first, we need to define special characters.
We decided not to add here dot or comma, since they are quite common.
Then we count with function *grepl* how many albums have special characters in their name and divide it by the number of distinct names of the albums

```{r}
special_characters <- "[!@#$%^&*()_+={}|:;\"<>?/~`\\]"

albums_with_special_chars <- df %>%
  filter(grepl(special_characters, album)) %>%
  select(album)

num_special_char_albums <- nrow(albums_with_special_chars)

total_albums <- nrow(df %>% distinct(album))

percentage_special_char_albums <- (num_special_char_albums / total_albums) * 100

num_special_char_albums
percentage_special_char_albums
```

### Albums that placed in top 25 in all three years

Here we choose albums that were ranked in top 25 in all three rankings

```{r}
df %>%
  filter(rank_2003 <= 25 & rank_2012 <= 25 & rank_2020 <= 25) %>%
  select(clean_name, album, genre, rank_2003, rank_2012, rank_2020)
```

## Plots

### Heatmap of number of albums from each genre in each rank

We are using previously calculated *number_of_albums_from_each_genre_in_each_rank* an vizualyzing it in a heatmap

```{r}
selected_colors <- c(stepped(20)[3], stepped(20)[11], stepped(20)[19])


ggplot(number_of_albums_from_each_genre_in_each_rank,
       aes(x = rank_year, y = genre, fill = album_count)) +
  
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradientn(
    colors = selected_colors,
    name = "Album Count",
    breaks = seq(0, max(number_of_albums_from_each_genre_in_each_rank$album_count) - 10, by = 10)
  ) +
  theme_minimal() +
  labs(title = "Number of Albums in Ranking",
       x = "Ranking Year",
       y = "Genre") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        legend.position = "right")
```

### Word cloud of the most used words

We are selecting names of the albums, tokenizing it, removing articles and other words that don't have meaning and then using function *worldcloud*, which does the vizualization.
The most common word is love.

```{r}
album_words <- df %>%
  select(album) %>% 
  tidytext::unnest_tokens(word, album) %>%  
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE)

album_words %>%
  arrange(desc(n)) %>% print()

selected_colors <- c(stepped(20)[9], stepped(20)[10], stepped(20)[13], stepped(20)[14], stepped(20)[17], stepped(20)[18])

wordcloud(
  words = album_words$word, 
  freq = album_words$n, 
  min.freq = 4,
  max.words = 50,
  random.order = FALSE,
  colors = selected_colors
)
```

### Average Age by Genre and Artist Gender in 2003

At first we need to filter the data, remove NA records from genre and artist gender.
After that we calculate average birth_year fo artists for groups and take exact birth year for solo artists and calculate age in 2003.
Then we group it by genre and artist gender and calculate average age for each gender and genre.
For some genres there aren't any female artists or groups that have both female and male members.
We can see that usually age of the female artists is lower in almost all genres.

```{r}
average_birthyear_by_genre_gender <- df %>%
  filter(!is.na(artist_gender) & !is.na(genre)) %>%
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>% 
  filter(!is.na(artist_gender) & !is.na(birth_year)) %>% 
  mutate(age = (2003 - birth_year)) %>%  
  group_by(genre, artist_gender) %>%
  summarise(average_birthyear = mean(birth_year, na.rm = TRUE), 
            average_age = mean(age, na.rm = TRUE),  
            .groups = "drop")

selected_colors <- c(stepped(20)[3], stepped(20)[15], stepped(20)[11])

ggplot(average_birthyear_by_genre_gender, aes(x = genre, y = average_age, color = artist_gender, group = artist_gender)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 4) +
  labs(title = "Average Age by Genre and Artist Gender in 2003",
       x = "Genre",
       y = "Average Age",
       color = "Artist Gender") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "gray80")) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, by = 10)) +
  scale_color_manual(values = selected_colors)
```

### Correlation Between Average Birth Year and Rank by Year and Genre

At first we need to filter the data, remove NA records from genre and artist gender.
After that we calculate average birth_year fo artists for groups and take exact birth year for solo artists.
We need to make longer data, so that we can implement the plot.
Names go to *rank_year* and values to *rank.* Then we need to filter again, remove NA records if exist for birth_year and rank.
We can see that a lot of artists were born in 1940-1950 years, which makes sense, since there was a "baby boom" after the II World War.

```{r}
debut_age_rank_data <- df %>%
  filter(!is.na(artist_gender) & !is.na(genre)) %>%
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>%
  pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>%
  filter(!is.na(birth_year) & !is.na(rank)) %>% 
  mutate(rank_year = gsub("rank_", "", rank_year))

plot <- ggplot(debut_age_rank_data, aes(x = rank, y = birth_year)) +
  geom_point(alpha = 0.7, size = 3, aes(color = genre)) +
  labs(title = "Correlation Between Average Birth Year and Rank by Year and Genre",
       x = "Rank",
       y = "Average Birth Year",
       color = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~rank_year, scales = "fixed") + 
  scale_color_manual(values = stepped(15)) +  # Use 15 colors from the stepped palette
  scale_y_continuous(limits = c(1910, 2000), 
                     breaks = seq(1910, 2000, by = 10)) +  
  scale_x_continuous(breaks = seq(0, 500, by = 100)) +
  theme(strip.text = element_text(size = 12, face = "bold"))

plot

ggsave("larger_graph.jpg", plot = plot, width = 16, height = 10, dpi = 300)
```

### Spotify Presence by Genre

We decided to see how many artists from different genres doesn't have Spotify (we thought that everyone has).
So we created additional column *has_spotify*, grouped by it and genre, summarized and added percentage.
Donut charts are very helpful in showing such information, so we decided to use them.

```{r}
spotify_analysis <- df %>%
  filter(!is.na(genre)) %>% 
  mutate(has_spotify = ifelse(!is.na(spotify_url), "Has Spotify", "No Spotify")) %>%
  group_by(genre, has_spotify) %>%
  summarise(album_count = n(), .groups = "drop") %>%
  mutate(percent = album_count / sum(album_count) * 100, .by = genre)


selected_colors <- c(stepped(20)[19], stepped(20)[20])
plot <- ggplot(spotify_analysis, aes(x = 2, y = percent, fill = has_spotify)) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y") +
  geom_text(data = spotify_analysis %>% filter(has_spotify == "No Spotify" & percent > 0),
            aes(label = paste0(round(percent, 1), "%"), y = percent / 2), 
            size = 4) +
  facet_wrap(~genre, nrow = 3) +
  labs(title = "Spotify Presence by Genre", fill = "Spotify Status") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  xlim(1.5, 2.5) + xlim(0.5, 2.5) +
  scale_fill_manual(values = selected_colors)
plot

ggsave("larger_graph.jpg", plot = plot, width = 16, height = 10, dpi = 300)
```

### Average Debut Age of Group vs. Solo Artists Across Genres

At first we need to filter the data, remove NA records from genre and artist gender.
After that we calculate average birth_year fo artists for groups and take exact birth year for solo artists, calculate debut_age, add new column Solo/Group, group by genre and artist type and find average debut age.
Usually, it's pretty similar, but in Latin average age for group is the highest one compared to all other genres and artist types.

```{r}
data <- df %>% 
  filter(!is.na(genre) & !is.na(artist_gender)) %>% 
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>% 
  mutate(debut_age = debut_album_release_year - birth_year) %>% 
  mutate(artist_type = ifelse(artist_member_count > 1, "Group", "Solo")) %>% 
  group_by(genre, artist_type) %>% 
  summarise(avg_debut_age = mean(debut_age, na.rm = TRUE), , .groups = "drop") 

selected_colors <- c(stepped(20)[16], stepped(20)[8])

plot <- ggplot(data, aes(x = genre, y = avg_debut_age, group = artist_type, color = artist_type)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 3) +  
  labs(title = "Average Debut Age of Group vs. Solo Artists Across Genres", 
       x = "Genre", y = "Average Debut Age", color = "Artist Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        panel.grid.minor = element_blank(), 
        legend.position = "top") +
  scale_y_continuous(limits = c(15, 50), 
                     breaks = seq(15, 50, by = 5)) +
  scale_color_manual(values = selected_colors)

plot 
```

### Album Rank Changes Over Time

At first we reshape the data and filter out missing values for rank and genre.
After that, we calculate how much each album’s rank changes over time by comparing its position to the previous year’s rank.
Finally, we create an animated plot showing the rank changes for each album, with colors representing the change and shapes showing the genre, using a custom stepped color palette for the ranks.

```{r}
data <- df %>%
  pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>% 
  filter(!is.na(rank) & !is.na(genre)) %>%
  mutate(rank_year = str_extract(rank_year, "\\d{4}")) %>% 
  mutate(rank_year = as.integer(rank_year)) %>% 
  arrange(rank_year, rank)

data <- data %>%
  group_by(album) %>%
  arrange(rank_year) %>%
  mutate(rank_change = rank - lag(rank, default = rank[1])) 


plot <- ggplot(data, aes(x = album, y = rank, color = rank_change, shape = genre)) +
  geom_point(size = 3) +
  scale_color_gradient2(low = stepped(24)[1], mid = stepped(24)[24], high = stepped(24)[13], midpoint = 0) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14)) + 
  labs(title = "Album Rank Changes Over Time: {frame_time}", 
       x = "Album", y = "Rank", color = "Rank Change", shape = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 14),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        ) +
  transition_time(rank_year) +  
  ease_aes('linear')  

animate(plot, fps = 10, duration = 8, width = 900, height = 600, renderer = gifski_renderer())
```

### Appearance of Emotional Words in Album Titles

Emotional words in album titles were counted and categorized as positive or negative.
The visualization shows that positive words appear more frequently, with "love" being the most common.

```{r}
positive_words <- c("Love", "Like", "Happy", "Joy", "Smile", "Dream", "Hope", "Bright", "Peace", "Sweet")
negative_words <- c("Hate", "Dislike", "Sad", "Cry", "Tears", "Dark", "Pain", "Lonely", "Broken", "Fear")

emotional_words <- c(positive_words, negative_words)

emotion_analysis <- df %>%
  separate_rows(album, sep = " ") %>%
  filter(album %in% emotional_words) %>%
  group_by(album) %>%
  summarise(appearance_count = n()) %>%
  arrange(desc(appearance_count))

emotion_analysis <- emotion_analysis %>%
  mutate(category = ifelse(album %in% positive_words, "Positive", "Negative"))

emotion_summary <- emotion_analysis %>%
  group_by(category) %>%
  summarise(total_appearances = sum(appearance_count))

emotion_plot <- ggplot(emotion_analysis, aes(x=reorder(album, -appearance_count), y=appearance_count, fill=category)) +
  geom_col() +
  theme_minimal() +
  scale_fill_manual(values = stepped()) +
  labs(title="Emotional Words in Album Titles", x="Word", y="Count") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
emotion_plot
```

### Artist Gender by Birth Decade

An approximated decade of birth was first calculated for each artist, then occurrences were counted and displayed.
Only in recent years more top female than top male artists were born.Also, most artists in the rankings were born in the '40s-'60s range.

```{r}
df <- df %>%
  mutate(birth_decade = floor(artist_birth_year_sum / artist_member_count / 10) * 10) %>%
  filter(!is.na(birth_decade))

gender_by_decade <- df %>%
  count(birth_decade, artist_gender) %>%
  spread(artist_gender, n, fill = 0)

g <- ggplot(gather(gender_by_decade, Gender, Count, -birth_decade), aes(x=birth_decade, y=Count, fill=Gender)) +
  geom_bar(stat='identity', position='dodge') +
  theme_minimal() +
  scale_fill_manual(values = stepped()) +
  scale_x_continuous(breaks = seq(min(1900), max(2000), by = 10)) +
  labs(title="Artist Gender Distribution by Birth Decade", x="Birth Decade", y="Count")

g
```

### Years Between Debut and Top Album in 2020

Average rank was calculated based on years between debut and 2020.
The plot suggests that artists with shorter careers achieve better rankings.

```{r}

df <- df %>% 
  mutate(years_between_top_2020 = release_year - debut_album_release_year,
         years_range = floor(years_between_top_2020 / 10) * 10,
         years_range_label = paste0(years_range, "-", years_range + 9)) %>%
  filter(!is.na(debut_album_release_year))

years_between_top_2020_range <- df %>%
  select(clean_name, years_between_top_2020, rank_2020, years_range_label) %>%
  filter(!is.na(rank_2020)) %>%
  group_by(years_range_label) %>%
  summarise(avg_rank = round(mean(rank_2020, na.rm = TRUE), 2)) %>%
  arrange(years_range_label)

ggplot(years_between_top_2020_range, aes(x = years_range_label, y = avg_rank, group = 1, color = years_range_label)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = stepped(length(unique(years_between_top_2020_range$years_range_label)))) +
  labs(title = "Average Rank vs. Years Between Debut and 2020 Hit",
       x = "Years Between Debut and 2020 Hit (Range)",
       y = "Average Rank",
       color = "Years Range") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Album Ranking Movements Over Time

This bar plot created by calculating average movement of ranks by release decade shows that newer albums tend to achieve higher places more easily.

```{r}
average_movement_by_decade <- df %>% 
  mutate(release_decade = floor(release_year / 10) * 10) %>%
  group_by(release_decade) %>%
  summarise(avg_movement = round(mean(differential, na.rm = TRUE), 2))

ggplot(average_movement_by_decade, aes(x = factor(release_decade), y = avg_movement, fill = factor(release_decade))) +
  geom_col() +
  scale_fill_manual(values = stepped(length(unique(average_movement_by_decade$release_decade)))) +
  theme_minimal() +
  labs(title = "Average Album Ranking Movement Over Time", x = "Release Decade", y = "Average Movement", fill = "Decade") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Average Length of Album Titles

An animated bar plot shows differences in album title length and rank divided by artist gender.
The result shows that for male artists the title length has little influence over rank, for female artists shorter titles are better, and for mixed groups - there is only a small difference in length between the top and bottom ranks.

```{r}
album_length <- df %>% 
  mutate(word_count = str_count(album, " ") + 1, char_count = nchar(album), rank_group = cut(rank_2020, breaks = seq(0, 500, 100))) %>% 
  drop_na(rank_2020)

animated_plot <- ggplot(album_length, aes(x = rank_group)) +
  geom_bar(aes(y = char_count, fill = "Character Count"), stat = "summary", fun = "mean", position = "dodge") +
  geom_bar(aes(y = word_count, fill = "Word Count"), stat = "summary", fun = "mean", position = "dodge") +
  scale_fill_manual(values = stepped(2)) +
  theme_minimal() +
  labs(title = "Album Title Length by Rank Group - {closest_state}", 
       x = "Rank Group", 
       y = "Average Count", 
       fill = "Legend") +
  transition_states(artist_gender)

animated_plot
```

### Relationship Between Debut Album Release Year, Release Year, and Highest Rank

A scatter plot was created to show the relationship between an artist's debut album release year and the number of years it took them to reach their top-ranked album.
The color gradient represents the highest rank achieved, with blue indicating better ranks and red indicating worse ranks.
The results suggest that artists debuting earlier generally took longer to reach their highest-ranked album, but there are exceptions across different time periods.

```{r}
debut_release_rank <- df %>% 
  mutate(highest_rank = pmin(rank_2003, rank_2012, rank_2020, na.rm = TRUE),
         year_difference = release_year - debut_album_release_year) %>% 
  select(debut_album_release_year, release_year, year_difference, highest_rank) %>% 
  arrange(highest_rank)

ggplot(debut_release_rank, aes(x = debut_album_release_year, y = year_difference, color = highest_rank)) +
  geom_point(alpha = 0.6) +
  scale_color_gradientn(colors = stepped(5)) +
  scale_x_continuous(breaks = seq(min(1930), max(2020), by = 10)) +
  theme_minimal() +
  labs(title = "Debut Year vs. Years to Top Album vs. Highest Rank",
       x = "Debut Album Release Year",
       y = "Difference in Years",
       color = "Highest Rank") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(10, 10, 10, 10)
  )
```

### Genre vs. Rank in a Particular Year

A heatmap was created to visualize the rankings of different genres over multiple years.
Higher rankings (closer to 1) are shown in blue, while lower rankings are in red.
As can be see, Reggae and Funk/Disco were present near the top of the ranking fo all three lists.

```{r}
df_long <- df %>% 
  select(genre, rank_2003, rank_2012, rank_2020) %>% 
  gather(year, rank, -genre) %>% 
  mutate(year = as.factor(gsub("rank_", "", year))) %>% 
  drop_na()

heatmap_plot <- ggplot(df_long, aes(x=year, y=genre, fill=rank)) +
  geom_tile() +
  scale_fill_gradientn(colors = stepped(5), trans = "reverse") +
  theme_minimal() +
  labs(title="Heatmap of Genres vs. Rank in Particular Years", x="Year", y="Genre", fill="Rank")

heatmap_plot
```

### Genre Proportions Over Decades

This plot shows how popularity of genres changes by decades based on their appearance in the rankings.
The results show that in the '70s Big Band/Jazz was most popular, with Hip-Hop/Rap taking its place in the recent decades.

```{r}
df$release_decade <- floor(df$release_year / 10) * 10

genre_proportion <- df %>% 
  count(release_decade, genre) %>% 
  group_by(release_decade) %>% 
  mutate(proportion = n / sum(n)) %>%
  filter(!is.na(genre))

proportion_plot <- ggplot(genre_proportion, aes(x=release_decade, y=proportion, color=genre, group=genre)) +
  geom_line(size=1) +
  geom_point(size=2) +
  scale_color_manual(values = stepped(length(unique(genre_proportion$genre)))) +
  scale_x_continuous(breaks = seq(min(genre_proportion$release_decade), max(genre_proportion$release_decade), by = 10)) +
  theme_minimal() +
  labs(title="Genre Proportions Over Decades",
       x="Release Decade",
       y="Proportion",
       color="Genre")

proportion_plot
```

### Name Length vs. Top Rank vs. Debut Year

This scatter plot examines the relationship between artist name length, debut year, and highest ranking, using color to indicate debut year trends.
As it can be seen, the average name length varies in the 9-13 character amount range and the longest names mostly belong in the '60s-'80s range.

```{r}
df <- df %>% 
  mutate(name_length = nchar(clean_name),
         highest_rank = pmin(rank_2003, rank_2012, rank_2020, na.rm = TRUE)) %>%
  drop_na(name_length, highest_rank, debut_album_release_year)

name_length_plot <- ggplot(df, aes(x=name_length, y=highest_rank, color=debut_album_release_year)) +
  geom_point(alpha=0.6) +
  scale_color_gradientn(colors = stepped(5)) +
  theme_minimal() +
  labs(title="Name Length vs. Top Rank vs. Debut Year", x="Name Length", y="Top Rank", color="Debut Year")

name_length_plot
```

### Lisa's garden based top 2 artists per genre based on album rankings

This circular plot based on tidytuesday's Lisa's Garden shows top 2 artists from each genre in a colorful way.

```{r}
data_clean <- df %>%
  select(clean_name, genre, rank_2003, rank_2012, rank_2020) %>%
  pivot_longer(cols = starts_with("rank_"), names_to = "year", values_to = "rank") %>%
  filter(!is.na(rank)) %>%
  filter(!is.na(genre)) %>%
  filter(!is.na(clean_name))

top_artists <- data_clean %>%
  group_by(genre, clean_name) %>%
  summarise(best_rank = min(rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(genre, best_rank) %>%
  group_by(genre) %>%
  slice_min(order_by = best_rank, n = 2, with_ties = FALSE)

edges_level0_1 <- top_artists %>%
  distinct(genre) %>%
  mutate(from = "Top Artists", to = genre)

edges_level1_2 <- top_artists %>%
  select(from = genre, to = clean_name)

edge_list <- bind_rows(edges_level0_1, edges_level1_2)

graph <- as_tbl_graph(edge_list, directed = TRUE) %>%
  activate(nodes) %>%
  mutate(genre = ifelse(name %in% top_artists$clean_name, 
                        top_artists$genre[match(name, top_artists$clean_name)], 
                        name))

color_palette_graph <- stepped(length(unique(top_artists$genre)) + 1)

p <- ggraph(graph, layout = 'dendrogram', circular = FALSE) + 
  geom_edge_diagonal() +
  geom_node_point(aes(color = factor(genre)), na.rm = TRUE) +
  geom_node_text(aes(label = name), repel = TRUE, na.rm = TRUE) +
  scale_color_manual(values = color_palette_graph) +
  theme_void()

node_positions <- ggplot_build(p)$data[[2]]

ggraph(graph, layout = 'dendrogram', circular = TRUE) +
  geom_edge_diagonal(color = "grey60") +
  geom_node_point(aes(color = factor(genre)), size = 3, na.rm = TRUE) +
  geom_node_text(aes(label = name), size = 3, repel = TRUE, na.rm = TRUE) +
  scale_color_manual(values = color_palette_graph) +
  theme_void() +
  ggtitle("Top 2 Artists per Genre")
```
